---
title: "Psychological Effectiveness in Sports Training"
author: "Steven Mou"
date: 2026-09-21
---


\

## Data Source:

This analysis uses the *Sports Training* dataset by
[VisionLangAI], obtained from Kaggle and licensed under
[CC BY 4.0](https://creativecommons.org/licenses/by/4.0/).

Source: https://www.kaggle.com/datasets/visionlangai/sports-training

## Intro
There exist many methods and intensities that athletes can use to train. From a general perspective, most people think the same two things when it comes to training: **Train hard, and Train smart**. 

But what does that really mean? Obviously, there are different sports and each sport has to train differrently. However, at the same time, each sport can take something away from another. Besides the sport, we need to consider the individuals playing the sport. Even if two players play the same sport, no two players will have the same psychological profile. 

Through an analysis of this dataset, we will analyze training methods, adaptation types, intensity level and other factors to answer the question: *Which psychological training methods are associated with better outcomes for athletes?*

## Loading the Data
```{r}
#| message: false
#| warning: false

# Loading the data and inspecting
library(tidyverse)

sports <- read_csv('data/psych_training_sports.csv')
glimpse(sports)
```
## Examining the Training Methods
```{r}
method_summary <- sports |>
    count(Training_Method, sort = TRUE) |>
    mutate(pct_total = n/sum(n) * 100)

method_summary
```
Here, we count all the available training methods recorded in this dataset and calculate their percentage. We can see that the most used training method for athletes is to set a goal for myself. However, this method is extremely comparable to self-talk and mindfulness as well, with all three hovering around 26% of the whole population. This makes sense as all three of these methods are what's taught in life even away from sports. People are taught since adolescence to set a goal for themselves, to think over their actions, and be aware of their actions and the subsequent results. What's interesting here, though, is that the lowest used method is cognitive restructuring: a method not very commonly heard of. 

## Summarizing Effectiveness by Method
```{r}
method_summary <- sports |>
    group_by(Training_Method) |>
    summarize(
        avg_motivation_gain = mean(Motivation_Gain, na.rm = TRUE),
        avg_confidence_gain = mean(Confidence_Gain, na.rm = TRUE),
        avg_stress_reduction = mean(Stress_Reduction_Percentage, na.rm = TRUE),
        avg_effectiveness = mean(Adaptation_Effectiveness_Score, na.rm = TRUE),
        .groups = 'drop'
    ) %>%
    left_join(method_summary, ., by='Training_Method')

method_summary |>
    print(width = Inf)

```

## Comparing Average Effectiveness
```{r}
#| label: fig-training-effectiveness
#| fig-cap: "Average adaptation effectiveness score by psychological training method."

ggplot(
  method_summary,
  aes(
    y = reorder(Training_Method, avg_effectiveness),
    x = avg_effectiveness
  )
) +
  geom_point(size = 4) +
  geom_text(
    aes(label = round(avg_effectiveness, 2)),
    nudge_y = 0.15
  ) +
  coord_cartesian(xlim = c(11.8, 12.2)) +
  labs(
    x = "Average adaptation effectiveness score",
    y = "Training method"
  ) +
  theme_minimal()
```

Even though some methods are preferred more than others, this figure above shows that the average adapatation effectiveness scores are very similar across all five training methods. 

Self-talk method had the highest average score at about 12.06 while mindfulness had the lowest average score at about 11.96. This only produces a spread between the highest and lowest score at 0.10 points, which suggests that the training method alone isn't very strongly associated with adaptation effectiveness in this dataset.

## Relationship between Adherence and Effectiveness
```{r}
tibble(
  measure = c("Pearson correlation", "Kendall's tau"),
  value = c(
    cor(
      sports$Training_Adherence_Rate,
      sports$Adaptation_Effectiveness_Score,
      use = "complete.obs",
      method = "pearson"
    ),
    cor(
      sports$Training_Adherence_Rate,
      sports$Adaptation_Effectiveness_Score,
      use = "complete.obs",
      method = "kendall"
    )
  )
)
```

With both Pearson's correlation (-0.009) and Kendall's tau (-0.005) nearing very closely to zero, we can see that there is essentially no linear nor monotonic association between adherence to training and the adaptation effectiveness score. This means that, within this dataset, higher adherence to training does not appear to correspond with higher or lower adaptation effectiveness.